{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-03-23 paper.tex
Start small tweaks

\begin{abstract}
User-centric programming research gave rise to a wide range of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming language and system paradigms.

We contribute Denicek, a computational substrate that reduces the complexity of
implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides three
primitive operations on series of edits, application of edits, merging of histories and conflict resolution.
We show that the three operations form the backbone of a direct implementation of a range of
programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple end-user data exploration environment.
The case study shows that the Denicek computational substrate provides an appealing basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

@@

A computational substrate defines the basic structures using which programs are constructed, how
the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a suitable program representation and operations for working with programs
provided by the substrate can eliminate much of the complexity of implementing interesting
programming experiences. For example, the reflective capabilities of Smalltalk make it easy
to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.

@@

What would be the ideal programming substrate for rich, interactive end-user programming systems?
In this paper, we aim to design a programming substrate that makes it easy to develop a range of
programming experiences \cite{myers-2006-eup}:

@@

>>>>

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming language and system.

We contribute Denicek, a computational substrate that makes it easier to
implement the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. It provides three
core operations on edit histories: (i) edit application, (ii) merging of histories and (iii) conflict resolution.
We show that these operations serve as fundamental building blocks for a direct implementation
of a range of programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of an interactive data science notebook.
The case study shows that the Denicek computational substrate provides a robust basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

@@

A computational substrate defines the structures from which programs are constructed, how
the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a suitable program representation can eliminate much of the complexity of implementing interesting
programming experiences. For example, the reflective capabilities of Smalltalk make it easy
to build rich debugging tools \cite{rauch-2019-babylonian} that are difficult to implement
for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.

@@

What would be the ideal programming substrate for rich, interactive end-user programming systems?
In this paper, we aim to design a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup} including:

@@

NICE THINGS IN DATNICEK
* merging of code edits (e.g. change parameter merges with adding a call to chain)
* remove cell to which there are refs - does the susbtrate catch this?

RELATED WORK

TIJS
https://homepages.cwi.nl/~storm/publications/icsews13live-p17-p-16201-preprint.pdf

via Pierre
https://pijul.org/manual/why_pijul.html

Geoffrey Litt
https://dl.acm.org/doi/pdf/10.1145/3426428.3426914
https://dl.acm.org/doi/pdf/10.1145/3397537.3397541

Programming By Example: Intelligence in Demonstrational Interfaces
https://cacm.acm.org/research/programming-by-example-intelligence-in-demonstrational-interfaces/

Rapido: Prototyping Interactive AR Experiences through
Programming by Demonstration
https://dl.acm.org/doi/pdf/10.1145/3472749.3474774

openrefine
- facet (group by) - then change the key to normalize
  (https://www.youtube.com/watch?v=yTJ6x6zEQmI 51:40)

Don't Step on My Toes: Resolving Editing Conflicts in Real-Time
Collaboration in Computational Notebooks
https://arxiv.org/pdf/2404.04695

AI assistants in notebook interfaces
https://dl.acm.org/doi/pdf/10.1145/3544548.3580940

Unravel - understand pipelines
https://dl.acm.org/doi/pdf/10.1145/3472749.3474744

WREX - nicest tool
https://dl.acm.org/doi/pdf/10.1145/3313831.3376442
